---
title: "Statistical Inference Course project, Part 2"
author: "oleg.reznychenko@gmail.com"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
documentclass: article
classoption: a4paper
font: Courier
fontsize: 11pt
geometry: margin=0.4in
---

#Objectives overview: 
Use confidence intervals and/or hypothesis tests to compare tooth growth by supp and dose.

#Exploratory analyisis:
Let's load the ToothGrowth data and explore some basic features.
```{r}
data(ToothGrowth)
tg <- ToothGrowth;
head(tg)
summary(tg)
t1<-rbind(c(class(tg$len), class(tg$supp), class(tg$dose)),
          c(length(unique(tg$len)), length(unique(tg$supp)), length(unique(tg$dose))));
colnames(t1) <- names(tg);
rownames(t1) <- c("Class of parameter", "Number of values");t1
```

#Data analysis
Parameters summary:      
 - **len** is continious quantitative data;     
 - **supp** is qualitative(categorical) data, indicate different states, the only possible operation is compare to equality;     
 - **dose** is discrete quantitative data with strong physical relations.     
 
From the scatterplot on Fig.1. we can see the **len** does not demonstrate clear dependency of **supp** - both groups are distributed widly and they are overlapped. Three groups of **dose** seems to be 'divided' much better. 

```{r, fig.width=8, fig.height=6}
pairs(tg, main="Fig.1. Scatterplot to illustrate data dependencies.", 
      lower.panel = panel.smooth, col="blue")
```
 
#Preliminary assumptions for further analysis:
We suppose:      
 - the all data groups are independent;    
 - we assume the constant variation across the groups;  
 - the $\alpha$-level is usual 5%;    
 - there are `r nrow(tg)` rows in the data set - the t-statistic should be used.     
 
The null-hypothesis $H_0$ will be an assumption the two groups mean values are equal.    
Function to calculate the 95% confidence interval to check the $H_0$:
```{r} 
confidence_interval95 <- function(x1,x2) {
  md <- mean(x1) - mean(x2);
  n1 <- length(x1); n2 <- length(x2);
  s1 <- sd(x1);   s2 <- sd(x2);
  semd <- sqrt(((n1-1)*s1^2 +(n2-1)*s2^2)/(n1+n2-2))*sqrt(1/n1+1/n2);
  md+c(-1,1)*qt(0.975, df=(n1+n2-2))*semd;
}
```

#Compare tooth growth by supp:

We create two samples, - for **supp** values ***OJ*** and ***VC***, - then draw a box plot with whiskers to illustrate the distributions of the **len** for each group (see Fig.2). 
```{r, fig.align='center', fig.width=5, fig.height=4}
tp.vc <- factor("VC", levels = levels(unique(tg$supp)))
tp.oj <- factor("OJ", levels = levels(unique(tg$supp)))
lnBySuppOJ <- tg[tg$supp == tp.oj, ]$len
lnBySuppVC <- tg[tg$supp == tp.vc, ]$len
boxplot(lnBySuppOJ,lnBySuppVC, 
        main="Fig.2. Boxplot of 'len' distributions by 'supp'.", 
        names=c("OJ", "VC"), col="lightgray")
rbind(c(mean(lnBySuppOJ),mean(lnBySuppVC)),
      c(sd(lnBySuppOJ), sd(lnBySuppVC)))
```
From the Fig.2. and the table above we see the both samples are wide spreaded and overlapped. Let's calculate the 95% confidence interval for $H_0$ hypothesis using function *confidence_interval95* and standard function t.test.
```{r}
confidence_interval95(lnBySuppOJ, lnBySuppVC)
```
\pagebreak
```{r}
ttest<-t.test(lnBySuppOJ,lnBySuppVC, paired=FALSE, var.equal = TRUE);ttest
```
#Conclusion of comparing tooth growth by supp:    
The 95% confidence interval for mean difference is [`r ttest$conf.int[1]`~`r ttest$conf.int[2]`], includes 0 value. The p-value is `r ttest$p.value`; it is small, but still greater then $\alpha$ (0.05). It means we **can not reject** $H_0$ hypothesis, i.e we can't say with full confidence the parameter **supp** affects on **len**. 

#Compare tooth growth by dose:
We create two samples, - for **dose** values > 1 and < 1 , - then draw a box plot with whiskers to illustrate the distributions of the **len** for each group (see Fig.3).
```{r, fig.align='center', fig.width=5, fig.height=4}
lnByDose2 <- tg[tg$dose > 1, ]$len
lnByDose05 <- tg[tg$dose < 1, ]$len
boxplot(lnByDose2, lnByDose05,
        main="Fig.3. Boxplot of 'len' distributions by 'dose'.", 
        names=c( "Dose = 2.0", "Dose = 0.5"), col="lightgray")
rbind(c(mean(lnByDose2),mean(lnByDose05)),
      c(sd(lnByDose2), sd(lnByDose05)))
```
Fig.3. ans the table above show the samples are more compact, - their standard deviations are smaller, - and less overlapped. Let's test the $H_0$ hypothesis.
```{r}
confidence_interval95(lnByDose2, lnByDose05)
ttest<-t.test(lnByDose2,lnByDose05, paired=FALSE, var.equal = TRUE);ttest
```
#Conclusion of comparing tooth growth by dose:   
The 95% confidence interval for mean difference is [`r ttest$conf.int[1]`~`r ttest$conf.int[2]`]. The p-value is `r round(ttest$p.value)`, it less then $\alpha$ (0.05). We **have to reject** $H_0$ hypothesis, i.e  for **dose**=2.0 value of **len** is definitly high then for **dose**=0.5. 


-----------------------------------------------------------------------
