---
title: "Statistical Inference Course project, Part 1"
author: "oleg.reznychenko@gmail.com"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
documentclass: article
classoption: a4paper
font: Courier
fontsize: 11pt
geometry: margin=0.4in
---

##Objectives overview: 
The objective is to investigate the distribution of averages of a number exponentials. The report contains of computer simulation of exponential distributed random variable, the sample mean and variance calculation and obtainig  the distribution of a large collection of averages of a number exponentials.  

##Simulation:
The exponential random variables were generated using command:
```{r eval=FALSE, tidy=FALSE}
exponential_randoms <- rexp(n, lambda)    # n- number of randoms, lamda-parameter of distibition
exprand_mean <- mean(exponential_randoms) # obtain the sample mean
exprand_var <- mean(exponential_randoms)  # obtain the sample variance
theory_mean <- 1/lmbd                     # calculate theoretical values of mean and variance
theory_var <- 1/lmbd^2
```

To calculate average of the sample of exponential randoms the next code was used:
```{r eval=FALSE, tidy=FALSE}
mns <- array(0, c(n))
for (i in 1:n) mns[i] = mean(rexp(sampleN, lmbd)); # generate 'sampleN' elements n times
samples_mns_mean <- mean(mns)
samples_mns_var <- var(mns)
salmpes_mns_theory_var <- theory_var/sampleN
```

```{r, fig.width=8, fig.height=4, echo=FALSE}
tseed<-as.numeric(Sys.time())
set.seed(tseed)
#set.seed(1423568583)


# function to perform statistical experiment
experiment1 <- function(lmbd, n) {
  # generate 'n' random variables
  rn_vals <- rexp(n, lmbd) 
  
  # calculate sample mean and variation
  mn_exprm <- mean(rn_vals);
  var_exprm <-var(rn_vals);
  sd_exprm <- sqrt(var_exprm);
    
  # calculate 'theoretical' mean and variation 
  mn_theor <- 1/lmbd
  var_theor <- 1/(lmbd^2)
  sd_theor <- sqrt(var_theor)

  list( rn_vals            = rn_vals,  
        mn_exprm           = mn_exprm,
        var_exprm          = var_exprm,
        sd_exprm           = sd_exprm,
        mn_theor           = mn_theor,
        var_theor          = var_theor,
        sd_theor           = sd_theor)
}

# function to perform statistical experiment with samples
experiment2 <- function(sampleN, lmbd, n) {
  
  # generate sample of 'sampleN' variables 'n' times
  # and store mean 
  mns <- array(0, c(n))
  for (i in 1:n) mns[i] = mean(rexp(sampleN, lmbd));   
  
  # calculate 'theoretical' mean and variation 
  mn_theor <- 1/lmbd
  var_theor <- 1/(lmbd^2)
  sd_theor <- sqrt(var_theor)

  # calculate both 'theoretical' and 'experimantal'
  # mean and variation for samples
  mn_mns <- mean(mns)
  var_mns <- var(mns)
  sd_mns <- sqrt(var_mns)
  var_mns_theor <- var_theor/sampleN
  sd_mns_theor <- sqrt(var_mns_theor)
  
  list( mns                = mns,
        mn_theor           = mn_theor,
        var_theor          = var_theor,
        sd_theor           = sd_theor,
        mn_mns             = mn_mns,
        var_mns            = var_mns,
        sd_mns             = sd_mns,
        var_mns_theor      = var_mns_theor,
        sd_mns_theor       = sd_mns_theor )
}

# plot histogram of 'full sample'
histogramExp <- function(figNo,lmbd, n, rn_vals, mn_exprm, mn_theor) {
  hist(rn_vals, main=paste('Fig.', format(figNo),'. Histogram of ', format(n,6),' random exponential'), col="palegreen", xlim=c(0,40), xlab=substitute(paste("Random value, PDF = ", lambda*e^{-lambda*x},", ",lambda, "=",lm), list(lm = lmbd)));
       
  abline(v=mn_exprm,col="red")
  abline(v=mn_theor, col='blue')
  
  SampleMeanStr <- paste("Sample mean = ", format(mn_exprm,digits=7));
  TheoryMeanStr <- paste("Theoretical mean =",format(mn_theor,digits=7));
  legend("topright", inset=.05, c(SampleMeanStr,TheoryMeanStr), fill = c("red", "blue"), horiz=FALSE)
}

# plot histogram of 'sample means experiment'
histogramMns <- function(figNo,lmbd, n, sampleN, mns, mn_mns, mn_theor, sd_mns_theor) {
  hist(mns, main = paste('Fig.',format(figNo),'. Histogram of ', format(sampleN),' random exponential \nmeans for ', format(n,6),' experiments'),
     xlab = "Value of the mean", ylim = c(0,0.6), col="ivory", freq=FALSE, breaks = 16);

  curve(dnorm(x, mean=mn_theor, sd=sd_mns_theor), add=TRUE, col="orange", lwd=2)
  lines(density(mns), col='green')

  abline(v=mn_mns,col="red")
  abline(v=mn_theor, col='blue')
  
  SampleMeanStr <- paste("Sample mean = ", format(mn_mns,digits=7));
  TheoryMeanStr <- paste("Theoretical mean =",format(mn_theor,digits=7));
  legend("topright", inset=.01, legend=c(SampleMeanStr,TheoryMeanStr,"Sample density","Theoretical density"), 
       fill = c("red", "blue", "green","orange"), horiz=FALSE)
}

# ------------------------------------------------------------------

n_list <- c(10000, 20000, 40000, 50000, 60000, 70000);
lmbd <- 0.2;

plotNum <- 1;
drawPlot <- array(0, c(length(n_list)));
drawPlot[1] <- TRUE;
drawPlot[(length(n_list))] <- TRUE;

# create empty data frame to store results
expResArrDf <- data.frame(NExp = numeric(),
                          MeanExp = numeric(), 
                          MeanTheor = numeric(),
                          VarExp = numeric(),
                          VarTheor = numeric(),
                          stringsAsFactors=FALSE);

colnames <- names(expResArrDf);
```

Simulation was run for **n** value in range [10000 ~ 70000], $\lambda=0.2$.
Resuts of the exponential random simulations are presented on histograms Fig.1. and Fig.2, and in Table 1.

```{r, fig.width=8, fig.height=5, echo=FALSE}

for (i in (1:length(n_list))){
  ES1 <- experiment1(lmbd,n_list[i]);
  
  # append result of the experiment
  expResArrDf <- rbind(expResArrDf,c(n_list[i], ES1$mn_exprm, ES1$mn_theor, ES1$var_exprm, ES1$var_theor));
    
  if (drawPlot[i]){
    histogramExp(plotNum, lmbd, n_list[i], ES1$rn_vals,ES1$mn_exprm, ES1$mn_theor )
    plotNum <- plotNum+1;
  }
}

names(expResArrDf) <- colnames;
```

$$Table\ 1.\ Exponential\ simulation\ experiment\ results.$$

```{r createtable1, results='asis', echo=FALSE}
cat(paste("Number of random values","Sample Mean","Theoretical Mean","Sample Var","Theoretical Var",sep = ' | '),paste(replicate(5,":---:"), collapse=' | '), sep="\n");
cat(apply(expResArrDf, 1, function(X) paste(X, collapse=" | ")), sep = "\n")
```

```{r, fig.width=8, fig.height=4, echo=FALSE}

sampleN_list <- c(20,30,40,50);
n_list <- c(10000, 20000, 40000);
lmbd <- 0.2;

drawPlot <- array(0, c(length(sampleN_list), length(n_list)));
drawPlot[1,1] <- TRUE;
drawPlot[4,3] <- TRUE;

# create empty data frame to store results
mnsResArrDf <- data.frame(NExp = numeric(),
                          NSample = numeric(),
                          MeanExp = numeric(), 
                          MeanTheor = numeric(),
                          VarExp = numeric(),
                          VarTheor = numeric(),
                          stringsAsFactors=FALSE);

colnames <- names(mnsResArrDf);
```

Simulation of samples average distribution was run for **n** value in range [10000 ~ 40000], **sample size** in range [20 ~ 50], $\lambda=0.2$.
Histograms and distribution of the exponential random samples average are presented on Fig.3., Fig.4. and in Table 2. 

```{r, fig.width=8, fig.height=5, echo=FALSE}
for (j in (1:length(n_list))) {
   for (i in (1:length(sampleN_list)))  {
      ES2 <- experiment2(sampleN_list[i],lmbd,n_list[j]);
      
      # append result of the experiment
      mnsResArrDf<- rbind(mnsResArrDf,c(n_list[j], sampleN_list[i], ES2$mn_mn, ES2$mn_theor, ES2$var_mns, ES2$var_mns_theor));
      
      if (drawPlot[i,j]){
        histogramMns(plotNum, lmbd, n_list[j], sampleN_list[i], ES2$mns,ES2$mn_mns, ES2$mn_theor, ES2$sd_mns_theor )
        plotNum <- plotNum+1;
      }
    }
}

names(mnsResArrDf) <- colnames;

```

\pagebreak

$$Table\ 2.\ Average\ of\ samples\ means\ experiment\ results.$$

```{r createtable2, results='asis', echo=FALSE}
cat(paste("Experiments","Sample","Sample Mean","Theor. Mean","Sample Var","Theor. Var",sep = ' | '),paste(replicate(6,":---"), collapse=' | '), sep="\n");
cat(apply(mnsResArrDf, 1, function(X) paste(X, collapse=" | ")), sep = "\n")

```

**Intermediate notes**:    
 - We see the sample average distribution is pretty close to normal one, the difference between obtained and theoretical values are about 0.1%;    
 - With these numbers of experiments we can't evaluate the sample variance behavior, it needs further investigation. 
 
 
Let's check whether the variation of the averages of sample is close to the theoretical one if number of experiments increases. 

The following code performs the experiment calculation:

```{r, fig.align='center', fig.height=6, fig.width=8}
diff_var <- NULL;
lambda = 0.2;
sample = 40;
var_theor <- (1/lambda)^2/sample;
for (n in seq(10000, 1000000, by=10000)) {
  re <- array(0,n);
  for (i in (1:n)) re[i] <- mean(rexp(sample,lambda));
  diff_var <- c(diff_var, var(re)-var_theor)
}
```

Result of experiment is shown on Fig.5

\pagebreak
```{r, echo=FALSE, fig.align='center', fig.height=6, fig.width=8}
plot(diff_var, main="Fig.5. Result of 40 exponential random\n average distribution", col="blue", type = 'b',
     xlab="Number of samples, x10000", ylab = "(sample var) - (theoretical var)")
abline(h=0, col="red")
```

**Conclusion**: As we can see the variance of sample average shows a tendency to approach to the theoretical value in accordance to CLT. 


 -----------------------------------------------------------------------
